# Building the predictor #

In this example BIOS RNAseq freeze1 data is used.

```{r, message=FALSE}
library(BIOSRutils)
data(rnaSeqData_freeze1_06032015BIOS)
```

The  RNAseq counts are normalized using `r Biocpkg("edgeR")` TMM-function.

Some experimentation showed that strong filtering improves the performance of the predictor!

```{r, message=FALSE}
library(edgeR)
library(GenomicFeatures)
d <- DGEList(counts=assays(rnaSeqData)$counts, remove.zeros=TRUE)
keep <- rowSums(d$counts > 100) > 0.99*ncol(d)
d <- d[keep,]
dim(d) #[1] 13778  2116
library(genefilter)
sds <- rowSds(d$counts)
keep <- sds > quantile(sds, prob=0.50)
sum(keep)
d <- d[keep,]
dim(d) #[1] 6889 2116
d <- calcNormFactors(d)
data <- cpm(d, log=TRUE) ##TMM CPM
data[1:5, 1:5]
```

Next cell percentages and covariates need to be extracted.

```{r}
cellPercentages <- as.data.frame(colData(rnaSeqData)[, c("Neut_Perc", "Lymph_Perc", "Mono_Perc", "Eos_Perc", "Baso_Perc")])
colnames(cellPercentages) <- c("Neutrophils", "Lymphocytes", "Monocytes", "Eosinophils", "Basophils")
cellPercentages <- apply(cellPercentages, 2, as.numeric)
head(cellPercentages)
```

```{r}
covariates <- as.data.frame(colData(rnaSeqData))
covariates <- covariates[, c("RNA_BloodSampling_Age", "Sex")]
colnames(covariates) <- c("Age", "Gender")
covariates <- apply(covariates, 2, as.numeric)
head(covariates)
```
Make sure all data object have the same id.

```{r}
rownames(covariates) <- rownames(cellPercentages) <-  colnames(data) <- colData(rnaSeqData)$uuid
head(covariates)
head(cellPercentages)
data[1:6, 1:6]
```
Find out which samples have complete cell percentages available.

```{r}
nas <- apply(cellPercentages, 1, function(x) any(is.na(x)))
table(nas)
```
And which are incomplete

```{r}
complete <- which(!nas)
head(cellPercentages[complete,])
```

split complete in train (2/3) and test (1/3) **stratified splitting of
the data might be better if sample size is small < 500? for example
using split.sample from caTools**

```{r}
trainId <- sample(complete, 2*length(complete)/3)
testId <- complete[!(complete %in% trainId)]
```

```{r}
dataTrain <- data[,trainId]
covarTrain <- covariates[trainId,]
cellPerTrain <- cellPercentages[trainId,]
```

Now build the predictor!

```{r, message=FALSE}
library(wbccPredictor)
pls.options(parallel=8) ##use 10 cores e.g. on the VM
predictor <- train(dataTrain, covarTrain, cellPerTrain, ncomp=250, validation = "CV", model = formula(cellPercentages ~ covariates + data), keep.model=TRUE)
```

The usual `r CRANpkg("pls")` functions can be used to inspect the
predictor-object. For example,

```{r, eval=FALSE}
summary(predictor)
cumsum(explvar(predictor))
```
and,

```{r}
barplot(cumsum(explvar(predictor)), las=2, ylab="Cumulative Variance Explained")
validationplot(predictor, val.type="R2", ncomp=1:50) ##select optimal number of components
validationplot(predictor, val.type="RMSEP", ncomp=1:250) ##select optimal number of components
validationplot(predictor, val.type="RMSEP", ncomp=1:50) ##select optimal number of components
```

Determine optimal number of pls-components e.g. 40 and save the
coefficients like this:

```{r}
RNAseqPredictor <- coef(predictor, ncomp = 40, intercept = TRUE)[,,1]
```

It is interesting to known which covariates have the highest
prediction value and store these as well.

```{r}
W <- predictor$loading.weights
ord <- order(abs(W[, 40]), decreasing=TRUE)
RNAseqTop <- gsub("data|covariates", "", rownames(W)[ord[1:1000]])
```

```{r}
save(RNAseqTop, file="../data/RNAseqTop.RData")
save(RNAseqPredictor, file="../data/RNAseqPredictor.RData")
```

# Some validation using the Test-set #

```{r, }
corrplot <- function(measured, predicted, xlab="predicted", ylab="measured", ...) {
    for(k in 1:ncol(measured)) {
        type <- colnames(measured)[k]
        pc <- signif(cor(predicted[,k], measured[,k]), 3)
        ic <- signif(icc(predicted[,k], measured[,k]), 3)
        plot(predicted[,k], measured[,k],
             main=paste(type, " (Prs: ", pc, ", ICC: ", ic, ")", sep=""),
             xlab=xlab, ylab=ylab, bty="n", pty="s", ...)
        abline(lm(measured[,k]~0+predicted[,k]), col=2, lty=1, lwd=2)
        abline(0, 1, col="grey", lty=1, lwd=2)
        grid()
    }
}

baplot <- function (x, y, regline = FALSE, la = c("log", "lin", "both"),
                    main = "", xlab = "Average", ylab = "Difference") {
    la <- match.arg(la)
    for (k in 1:ncol(x))
        wbccPredictor:::.baplot(x[, k], y[, k],
                                regline = regline,
                                la = la, main = paste0(colnames(x)[k]),
                                xlab = xlab, ylab = ylab)

}
```

This is no surprise!

```{r}
predictedRNATrain <- predictor$fitted.values[,,40]
op <- par(mfcol=c(3,2))
corrplot(cellPerTrain, predictedRNATrain, xlab="predicted", ylab="measured")
par(op)
```
Now let predict cell composition for the test data and plot.

```{r}
dataTest <- data[,testId]
covarTest <- covariates[testId,]
cellPerTest <- cellPercentages[testId,]
predictedRNATest <- prediction(predictor, dataTest, covarTest, ncomp=40, transformation=function(x) x)
op <- par(mfcol=c(3,2))
corrplot(cellPerTest, predictedRNATest, xlab="measured (%)", ylab="predicted (%)")
par(op)
op <- par(mfcol=c(3,2))
baplot(cellPerTest, predictedRNATest, la="lin", xlab="Average: meas. - pred. (%)", ylab="Difference: meas. - pred. (%)", regline=FALSE)
par(op)
```
